\chapter{Environment}
\label{chapter:environment}

TODO: expand on the business model. 

The client for this program is in the house maintenance business. They deliver and install various specific features to a house.

The distances to the customers vary greatly and can range from hundreds of kilometres to being in the same neighbourhood.  

When planning routes for the technicians, one has to take into account the requirements for each job and the limits of the workers and vehicles. There are also date and time constraints involved. The jobs have to be scheduled within 4-week windows determined by customers' preferences. Likewise, there may be specific hours of day during which the job must be done, if the customer needs to be present to let the technicians inside, for example. 

The required building materials must fit into the vehicle and the total hours of the working day typically should not exceed 8 hours. 

Different technicians have different skills, and some may perform faster than others. Some jobs can be done by a single technician, while others may require multiple people to do the job. 

The type of the house also plays a role. A detached house has different set of challenges than an apartment building. The size of the house also affects on the difficulty as well as the equipment and time requirements of the job.

Currently planning routes for the technicians is done by hand by an employee dedicated to the task. Due to the mechanical, repetitive and complex nature of the task, it is a suitable target for a computer algorithm to solve.

Because the vehicle routing problem is typically associated to solving cargo deliveries or pickups, applying it here requires some customisation to take the various constraints and requirements into account. 

\section{Input data source}
\label{subsection:dataconversion}

The input data for the program comes from the client company's sales data. The data contains the following information:

\begin{itemize}  
\item The address of the customer
\item The type of the house
\item Types and number of tasks involved in the job 
\item The items to be installed to the house or the items previously installed in case of a service visit
\item Equipment required for the job.
\item The time window for the job
\item Minimum number of technicians required.
\item Estimates of the technicians' performances. How fast they are at the job.
\end{itemize}

NOTE: number of technicians might always be one. This needs checking. Overall this is a very crude sketch of the subsection. 

To prevent the algorithm from becoming overly complex, the time and resource requirements of a single job should be simplified as much as possible. The end result should be the following parameters per job target:

\begin{itemize}
\item Vehicle capacity requirements of the installation items and other equipment.
\item Minimum number of technicians required at the job.
\item Total time requirement for the technicians assuming that the number of technicians is the minimum number.  
\end{itemize}